scTOP: physics-inspired order parameters for cellular identification and visualization

### 2023-07-21 manuscript/scTOP response to reviewers/review_response.tex

\documentclass[12pt,american]{scrartcl}
\usepackage{babel}
\usepackage[babel]{microtype}
\usepackage[babel]{csquotes}
\usepackage[margin=0.75in]{geometry}

\usepackage[journal={Development},
			manuscript={DEVELOP/2023/201873},
			editor={Paul Francois}]{reviewresponse}

\usepackage[T1]{fontenc}
\usepackage{lmodern}
%\usepackage{newcent}
%\usepackage[scaled]{beramono}

\usepackage[backend=biber,style=ieee,dashed=false,url=false,isbn=false,defernumbers=true,refsection=section]{biblatex}
\bibliography{literature.bib}

\usepackage{hyperref}

\title{scTOP: physics-inspired order parameters for cellular identification and visualization}
\author{Maria Yampolskaya, Michael J Herriges, Laertis Ikonomou, Darrell N Kotton, and Pankaj Mehta}

\begin{document}
% \maketitle

% % Cover Letter
% \input{cover_letter.tex}

% Reviewer 1
\reviewer

\begin{revcomment}
	the important fact that the method is linear is mentioned but its implication is not really discussed. Many people will think that linear method = PCA (for instance), and it should be mentioned that the linear computation used is different from unsupervised linear methods such as PCA and SVD ( because of the reference set). Also the method could prove useful way beyond single cell data because of this (for instance many people use PCA for morphometrics analysis, and I could see how having a reference dataset to perform computation would help there).
\end{revcomment}
\begin{revresponse}
	We agree that this method is general enough to be applied to other contexts. We have added the following section to the Supplementary Information to explain how scTOP relates to other linear methods:
	\begin{changes}
		\subsection*{Comparison to other linear methods}
          scTOP is mathematically and conceptually distinct from other linear methods such as Principal Component Analysis (PCA). PCA takes input data and finds the vectors in feature space along which the most sample-to-sample variation occurs. PCA is unsupervised and assumes nothing is known about the input data. The principal components need to be examined to determine their biological relevance. The major difference with scTOP is that scTOP takes input data and a reference basis of known cell types. The input data is decomposed onto the vectors defined by the reference basis instead of vectors corresponding to the most sample-to-sample variation. With scTOP, the resulting decomposition has a direct interpretation: alignment with each of the known cell types within the reference basis. It’s not possible to find the decomposition onto known cell types with PCA, because PCA doesn’t have the extra information about pre-defined cell type expression profiles.

          To summarize, scTOP and PCA both transform the sample data to a different coordinate system, but there are major differences between the spaces that these methods project onto. The following table breaks down the differences between these spaces.

          \begin{tabular}{ |p{2.5in}|p{2.5in}|} 
             \hline
             scTOP space & PCA space\\ 
             \hline
             \hline
             Defined by an input reference basis of known cell types & Calculated by PCA by finding the orthogonal vectors (principal components) that best explain the variation in data  \\ 
             \hline
             Can be non-orthogonal (e.g. can have highly-correlated cell types) & Orthogonal by definition  \\ 
             \hline
             Dimensions have clear biological interpretation: alignment with cell types & Dimensions need to be closely analyzed to be interpreted \\
             \hline
         \end{tabular}
	\end{changes}
\end{revresponse} \label{PCA}

\begin{revcomment}
	I see what the authors mean when they say it is important that the algorithms should give deterministic results, but in some cases, well controlled stochastic results could be necessary/important. I was thinking of MCMC based models, or similarly, instead of computing a deterministic order parameter, one could imagine algorithms giving meaningful probability scores while not being fully deterministic. In the end, the choice of the reference base certainly adds a quenched disorder component. Maybe the authors can elaborate a bit on this deterministic/stochastic issue because I suspect some people might not necessarily see the deterministic aspect as a strength 
\end{revcomment}
\begin{revresponse}
    We agree that stochastic results can be valuable. Rather than focus on the deterministic aspect of scTOP, we have edited the text to emphasize the importance of reproducibility rather than determinism. As you say, stochastic likelihood scores can also be useful; the important thing is the ability to make comparisons across analyses and data sets. We have changed the manuscript's mentions of determinism and stochasticity. For example, instead of ``Second, the algorithm should be deterministic, further enhancing reproducibility,'' the text now states:

    \begin{changes}
        Second, the results of the algorithm should be reproducible, so that direct comparison between data sets is possible.
    \end{changes}
\end{revresponse}

\begin{revcomment}
    the ‘mathematical details’ part is very condensed. It is certainly 
very readable for a physicist, but I do not think developmental 
biologists (even the ones interested in ‘modelling’) will be able to 
understand statements like ‘decorrelated versions of the conventional 
spin-glass magnetization’. I would recommend a short crash course for 
biologists on what an order parameter is, since it is the crucial 
concept to define scores. I like equations but I am not even sure they 
should be in the main text, I could for instance imagine a version of 
the paper where those concepts are explained in an intuitive/graphical 
way. At the very least more ‘intuitive’ details should be given to 
biologists.
\end{revcomment}
\begin{revresponse}
    As per the formatting guidelines for Development, we have moved the mathematical details to the ``Materials and Methods'' section after ``Discussion.'' We have also expanded the introduction section to give a more detailed explanation of what an order parameter is (the additions are in \textcolor{blue}{blue}).

    \begin{changes}
        Here, we present a new physics-inspired algorithm, single-cell Type Order Parameters (scTOP), that has all of these desired properties. The algorithm is based on the statistical physics concept of an order parameter -- a coarse-grained quantity that can distinguish between phases of matter. \textcolor{blue}{An order parameter takes on different values depending on the phase of the system. Typically, in a disordered phase, the corresponding order parameter is zero, and in an ordered phase it is nonzero.} Arguably the most famous example of an order parameter is magnetization in a ferromagnetic system, which is a collection of spins \textcolor{blue}{(in the simplest case, spins can be ``up'' or ``down'')} which align with one another. In a ferromagnet, the magnetization measures the alignment of spins and can be used to distinguish magnetic and paramagnetic phases. \textcolor{blue}{Below a particular temperature (known as the ``critical temperature''), the spins in the ferromagnet align with one another. In a simple model, the order parameter (magnetization, which measures the average spin) is approximately one if all the spins are up and approximately negative one if all the spins are down. When the temperature exceeds the critical temperature, the system changes from the magnetic phase to the paramagnetic phase, and the spins do not align; instead, they are randomly up or down, and the order parameter is approximately zero.}

        It is straightforward to generalize the idea of magnetization to more complex settings such as spin glasses and attractor neural networks. In the context of attractor-network epigenetic landscape models of cell identity, this takes the form of generalized magnetizations that measure how aligned a gene expression state is with each of the attractor basins for different cell fates. \textcolor{blue}{Whereas magnetization for the ferromagnetic system measured whether the system was in a magnetic or paramagnetic phase, these generalized magnetizations measure which phase a gene regulatory network is in (where the different phases correspond to different cell types).} These generalized magnetizations serve as natural order parameters for cellular identity and can be calculated directly from data. The scTOP algorithm builds on previous work by developing methods for integrating scRNA-seq data, greatly expanding the power and utility of this approach. 
    \end{changes}
\end{revresponse}

\begin{revcomment}
    in physics magnetization and spins could range from -1 to +1, here the
 order parameter is always positive (and lower than 1) as far as I can 
see. why ? In particular it is not entirely clear to me how things were 
put back between 0 and 1. Also, the maximum value of the order parameter
 computed from data seems to be around 0.5, can the authors comment on 
the magnitudes computed by their algorithm and how to interpet them ?
\end{revcomment}
\begin{revresponse}
    scRNA-seq data is very sparse, so it’s not possible to reach perfect alignment (score of +1). The reference basis contains aggregate profiles, where expressions have been averaged across cells of the same type. This removes many of the zeroes, but the single-cell samples still have many zeroes, resulting in misalignment. When samples are averaged in order to find aggregate scores, it’s possible to reach higher scores thanks to the reduction in zeroes. The figure with the aggregate samples from the Mouse Cell Atlas compared to organs from Tabula Muris illustrates these higher scores. It also shows a few scores that are negative.

    We have also added the following section to the Supplementary Information:
    \begin{changes}
        \subsection*{Interpreting scTOP scores}
        In theory, the attractor-network order parameters we have adapted for gene expression data range between negative one and positive one. An attractor state order parameter with a value of positive one means the system is perfectly aligned with that particular attractor state, and negative one means it is the exact opposite of that attractor state. Zero means that it is neither correlated nor anti-correlated with that attractor state.

        In practice, scTOP scores generally do not reach positive one. scRNA-seq data contains significant numbers of dropouts, and these zeroes decrease the magnitude of the scTOP score. When an aggregate score is taken, the effect of dropouts is decreased and the scTOP score increases. This is important to keep in mind when interpreting scTOP scores; any score above 0.1 is notable, and since the data is so noisy even the best-case score may be well below 1. It's helpful to have a control population to compare with, to have a measure of the range of best-case scTOP scores. This is also why it's best to compare distributions of scores rather than scores for individual cells, since noise plays a non-negligible role.

        scTOP scores are rarely negative. Negative values for scTOP scores correspond to cases where the sample gene expression profile is the opposite of the reference profile. For example, a score of -1 would indicate that all the genes which are highly-expressed in the reference are lowly-expressed in the sample, and vice versa. There are cases where the scores are low, negative values, indicating that the sample does not align well with the reference and that some genes which are highly- (lowly-) expressed in the reference are lowly- (highly-) expressed in the sample. From the data, it appears that it’s unlikely for gene networks to contain cell types which are opposites of one another. One possible reason is that some genes, like housekeeping genes, are similarly expressed across all cell types. Also, it’s possible that cells which expressed the exact opposite genes of any given cell type would not be able to survive.
    \end{changes}
\end{revresponse}

\begin{revcomment}
    when commenting on $S^\perp$, it is qualified as ‘biological noise’, but
 then reference to cell cycle genes, etc… is made. I can imagine genes 
related to, say, oscillating genes in the segmentation clocks, which 
would definitely be thrown away by the method (because they do not 
correspond to actual cell fates) but are definitely not developmental 
noise either. Maybe some comment should be added on situations like 
this, where gene dynamics not clearly associated with ‘fates’ (as 
defined by humans in their reference dataset) can not really be 
detected/visualized with this method. Along the same lines, it could be 
useful to show what happens with ‘virtual’ experiments where some 
reference fates are removed to contrast them with the situation with all fates
\end{revcomment}
\begin{revresponse}
    We agree that one of the limitations of scTOP is that it gives only information on cell differentiation. It cannot capture the dynamics of gene expression beyond the changing of alignment with cell identities defined by the reference basis. To clarify this point, we have edited the $S^{\perp}$ paragraph:

    \begin{changes}
        Biologically, $S^{\perp}$ contains information about processes that affect gene expression but are not associated with cell identity, such as the cell cycle. We refer to the $a^{\mu}$ as the scTOP scores. Each component of this vector measures the projection of the cell with gene expression profile $S_i$ onto the $\mu$ - the cell type in the reference basis. These scores can be used to accurately classify cell identity and provide a natural visualization of gene expression in the space of possible cell fates. Gene regulatory dynamics that are unrelated to transitions between cell types in the reference basis would not be captured by these scores; instead, these would contribute to $S^{\perp}$. As a result, these fate-unrelated dynamics cannot be visualized using scTOP.
    \end{changes}
\end{revresponse}

\begin{revcomment}
The authors only show ‘2D’ representation, for two order parameters. 
Is this because they only see binary choices in the data? Could we see 
in 3D a sequence of two binary choices for instance ? Is there anything 
biological to tell there?
\end{revcomment}
\begin{revresponse}
    It can be difficult to distinguish the positions of points in a 3D scatter plot. We have included figure \ref{3D plots} here to illustrate this. In this figure, we plot a clonal family from the hematopoietic data set in the main text. This is a case where multiple cell identities are of interest, but it is difficult to discern their relationships from a 3D plot. In cases where multiple fates are of interest, in binary choices or otherwise, we use pair-plots to show multiple 2D projections simultaneously.

    \begin{figure}
	\centering
    \includegraphics[scale=0.3]{figs/hem 3d neutrophil-monocyte.png}
	\caption{An example of a three-dimensional representation of scTOP scores, in the case of hematopoietic differentiation. Even with the inclusion of "shadows" (the corresponding 2D projections on the boundaries) it is difficult to determine the relationships between cells, which is why we use pair-plots in the main text instead. }
	\label{3D plots}
    \end{figure}
\end{revresponse}

\begin{revcomment}
    Table I : it is a bit unclear to me how the method was validated when 
Query data are used on the same reference set. I was expecting something
 classic like an initial separation between training/validating 
ensembles, etc… But it rather seems from the Supplement that the 
training set essentially was based on ‘well-chosen’ cells or something 
like this. It could be useful to give more details in the main text.
\end{revcomment}
\begin{revresponse}
    We do have the initial separation between training and test data. For clarity, we have changed the table column names to ``Test data source'' and ``Reference data source,'' and we have added the following line to the main text:
    \begin{changes}
        For cases where the training (or ``reference'') data source is the same as the test data source, we set aside a randomly-selected subset of cells for the training set and do not use them in the test data.
    \end{changes}

    The discussion of curating data sets in the SI refers to the cell types, rather than the cells, that were chosen for the reference basis. In other words, we discard cell types that have fewer than 100 corresponding cells. We do this because scRNA-seq data has a lot of dropouts, so we need to average across many cells to get a reasonable reference basis. We want to make sure the reference gene expression profiles are reflective of the true gene expression profiles of the cell types; making sure there are enough training data points per cell type is what we mean by having "well-sampled" reference profiles.

    To demonstrate the effect of not having enough cells per cell type, we have included figure \ref{sampling}. In this figure, we vary the number of cells sampled for the reference cell type profile, and find the corresponding scTOP score for the query data. With very few cells, the score is low since the reference type is poorly sampled. As the number of cells increase, the score plateaus. Ideally, we would have thousands of cells for every cell type, but unfortunately many single-cell data sets do not have so many cells per cell type.

    \begin{figure}
	\centering
    \includegraphics[scale=0.4]{figs/sampling.png}
	\caption{A plot of adult mouse lung stromal cells from the Mouse Cell Atlas projected onto a reference profile calculated from mesenchymal cells from the Atlas of Lung Development \cite{negretti2021single}. The x-axis indicates the number of cells sampled for the reference cell type, and the y-axis indicates the scTOP score. }
	\label{sampling}
    \end{figure}
\end{revresponse}

\begin{revcomment}
along the same line, all results depend on the quality of the 
reference set. Is there a way to somehow benchmark it ? For instance, I 
could imagine that a bad reference set could give a lot of false 
negatives or false positives, how would we see this ? (maybe by 
comparing with some unsupervised method ?)
\end{revcomment}
\begin{revresponse}
    In order to benchmark the quality of the reference bases, we measured the accuracy scores of control samples scored with those bases. In the manuscript, we provide examples of comparing the accuracy metrics with unsupervised methods from Abdelaal et al. In particular we compare the F1 score for scTOP using the CellBench and Allen Mouse Brain atlas with the F1 scores achieved for the same data sets using a variety of automatic cell identification algorithms, including unsupervised methods. The F1 score is a metric that is calculated from the true positives, false positives, and false negatives.

    Similar algorithms are often black boxes that are difficult to troubleshoot, but since scTOP simply uses linear algebra, it is possible to determine which genes are contributing to the scTOP scores. If a reference basis is giving a dubious result, it's possible to check if the top-contributing genes correspond to relevant biological processes. If not, this indicates that the reference basis may be poorly constructed, whether due to not having enough cells per cell type or not having distinct enough cell types.
\end{revresponse}

\begin{revcomment}
    The paper reads a bit long to me (12 figures, please check ). Maybe 
some figures with very similar messages/methods could be combined (e.g. 
4,5,6) and the writing could be a bit tightened there as well. There is a
 lot of ‘empty space’ in some figures (eg 9, 12) which could be 
tightened
\end{revcomment}
\begin{revresponse}
    In addition to ensuring that all figures match the formatting guidelines, we have made the following changes:
    \begin{enumerate}
        \item We have adjusted the layout and removed whitespace from figures 1, 5, and 6
        \item We have removed redundant panels (or moved them to the Supplementary Information), as in figures 5 and 9
        \item We have moved figure 8 to the Supplementary Information
        \item We have combined three figures into the panels of figure 4, and tightened the writing accordingly
    \end{enumerate}
\end{revresponse}

\begin{revcomment}
    1st sentence : « humans and other mammals… » I guess other animals too
\end{revcomment}
\begin{revresponse}
    We have fixed this line to say "humans and other animals."
\end{revresponse}

% Reviewer 2
\reviewer

\begin{revcomment}
	What is the relationship of scTOP to PCA? Presumably one could do PCA with C principal components for C distinct cell types and construct a new coordinate system as the average vector to each celltype. How should one think about this wrt to scTOP’s basis vectors? Can you perhaps provide a comparison of how the two methods fair when thinking about cell types or intermediates and even lineages
\end{revcomment}
\begin{revresponse}
 Although scTOP and PCA are both linear methods, there are significant differences between them. The principal components computed by PCA are the orthogonal dimensions that describe the most variance in the data. For example, the first principal component is the dimension along which the data varies the most. Taking C principal components would not provide information about alignment to the C cell types; instead, this coordinate system would give information about the top C dimensions along which there is the most variance. For example, if the input data corresponded to cells that were all of cell type A, the principal components would give information about how the gene expression varies between these cells, regardless of cell type. On the other hand, scTOP would identify all these cells as cell type A.

 We have added the following section to the Supplementary Information to explain how scTOP relates to other linear methods:
	\begin{changes}
		\subsection*{Comparison to other linear methods}
          scTOP is mathematically and conceptually distinct from other linear methods such as Principal Component Analysis (PCA). PCA takes input data and finds the vectors in feature space along which the most sample-to-sample variation occurs. PCA is unsupervised and assumes nothing is known about the input data. The principal components need to be examined to determine their biological relevance. The major difference with scTOP is that scTOP takes input data and a reference basis of known cell types. The input data is decomposed onto the vectors defined by the reference basis instead of vectors corresponding to the most sample-to-sample variation. With scTOP, the resulting decomposition has a direct interpretation: alignment with each of the known cell types within the reference basis. It’s not possible to find the decomposition onto known cell types with PCA, because PCA doesn’t have the extra information about pre-defined cell type expression profiles.

          To summarize, scTOP and PCA both transform the sample data to a different coordinate system, but there are major differences between the spaces that these methods project onto. The following table breaks down the differences between these spaces.

### 2023-07-21 manuscript/scTOP (Company of Biologists LaTeX template)/SI.tex
added revised manuscript

	\documentclass[aps,superscriptaddress, notitlepage,longbibliography]{revtex4-1}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{physics}
\usepackage[pdftex]{graphicx}
\usepackage{xcolor}
\usepackage{hyperref}
\setlength {\marginparwidth }{2cm}
\hypersetup{
    colorlinks,
    citecolor=blue,
    filecolor=black,
    linkcolor=red,
    urlcolor=blue
}

@@

\title{ scTOP: physics-inspired order parameters for cellular identification and visualization}

\author{Maria Yampolskaya}
\email{mariay@bu.edu}
\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
\author{Michael Herriges}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{The Pulmonary Center and Department of Medicine, Boston University School of Medicine, Boston, MA, USA}
\author{Laertis Ikonomou}
\affiliation{Department of Oral Biology. University at Buffalo, The State University of New York, Buffalo, NY, USA}
\affiliation{Division of Pulmonary, Critical Care and Sleep Medicine, Department of Medicine, University at Buffalo, The State University of New York, Buffalo, NY, USA}
\author{Darrell Kotton}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{The Pulmonary Center and Department of Medicine, Boston University School of Medicine, Boston, MA, USA}
\author{Pankaj Mehta}
\email{pankajm@bu.edu}
\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{Faculty of Computing and Data Science, Boston University, Boston, MA 02215, USA}
\affiliation{Biological Design Center, Boston University, Boston, MA 02215, USA}

\begin{abstract}
Advances in single-cell RNA-sequencing (scRNA-seq) provide an unprecedented window into cellular identity. The increasing abundance of data requires new theoretical and computational frameworks for understanding cell fate determination, accurately classifying cell fates from expression data, and integrating knowledge from cell atlases. Here, we present single-cell Type Order Parameters (scTOP): a statistical-physics-inspired approach for constructing ``order parameters'' for cell fate given a reference basis of cell types. scTOP can quickly and accurately classify cells at a single-cell resolution, generate interpretable visualizations of developmental trajectories, and assess the fidelity of engineered cells. Importantly, scTOP does this without using feature selection, statistical fitting, or dimensional reduction (e.g., UMAP, PCA, etc.). We illustrate the power of scTOP utilizing a wide variety of human and mouse datasets (both {\it in vivo} and {\it in vitro}). By reanalyzing mouse lung alveolar development data, we characterize a transient perinatal hybrid alveolar type 1/alveolar type 2 (AT1/AT2) cell population that disappears by 15 days post-birth and show that it is transcriptionally distinct from previously identified adult AT2-to-AT1 transitional cell types. Visualizations of lineage tracing data on hematopoiesis using scTOP confirm that a single clone can give rise to as many as three distinct differentiated cell types. We also show how scTOP can quantitatively assess the transcriptional similarity between endogenous and transplanted cells in the context of murine pulmonary cell transplantation. Finally, we provide an easy-to-use Python implementation of scTOP. Our results suggest that physics-inspired order parameters can be an important tool for understanding development and characterizing engineered cells.
\end{abstract}

@@

\section{Introduction}
Humans and other mammals have many specialized cells that arise through a complex developmental process  \cite{zeng_what_2022}. At the blastocyst/epiblast stages of embroynic development, cells are pluripotent and possess the flexibility to develop into fates from all three germ layers \cite{Wolpert_2019}. However, as embryos mature, cellular identity becomes less plastic and more specified through a complex spatiotemporal process involving signaling and other environmental cues. The questions of how and why cells end up in one fate or another are fundamental to developmental biology \cite{stephenson_intercellular_2012, zhu2020principles, rand2021geometry}. More practically, understanding the process of cell fate specification is crucial for developing directed differentiation protocols for engineering cells \emph{in vitro}, with important clinical applications for human health and disease \cite{rossant_impact_2011, rowe2019induced, basil2020cellular}.

One powerful set of techniques for characterizing cellular phenotype is transcriptome-wide measurement using single-cell RNA sequencing (scRNA-seq). Typical scRNA-seq experiments involve thousands of genes and dozens to millions of cells. For this reason, this data is intrinsically high-dimensional and requires specialized computational tools for analysis and interpretation \cite{zhu2020principles, setty2019characterization, schiebinger2019optimal}. This task is especially difficult because the technical challenges of sequencing RNA from individual cells result in data that is noisy and sparse \cite{hicks2018missing, lahnemann_eleven_2020, argelaguet2021computational}. Genes with relatively low expression are often recorded as not being expressed at all, resulting in zero counts called dropouts. Experiment-to-experiment differences in sequencing methods, platforms, and other difficult-to-standardize conditions result in technical variations in scRNA-seq measurements called batch effects. These challenges merit careful consideration of how to normalize and analyze the data in a way that permits the study of biological variation without conflation by technical variation.

scRNA-seq is often used for cellular identification. The most frequently used workflow involves applying dimensional reduction (such as Uniform Manifold Approximation and Projection (UMAP) \cite{mcinnes_umap_2018} or SPRING \cite{weinreb_spring_2018}) to visualize the data in two dimensions, then using unsupervised clustering to group individual cells. These groups are then often manually annotated according to expression levels of known marker genes. Machine learning classifiers such as support vector machines are also commonly used to identify cells after training with labeled data \cite{abdelaal_comparison_2019}. Another common analysis task is trajectory inference, where the goal is to infer developmental trajectories from static time shots (reviewed in \cite{saelens2019comparison,wagner2020lineage}). Trajectory inference arranges cells on a low-dimensional manifold and assigns pseudo-times according to transcriptomic similarity. There are also methods such as partition-based graph abstraction \cite{wolf_paga_2019} that purport to combine clustering and trajectory inference to simultaneously calculate pseudo-times and provide discrete labels at varying resolutions. RNA-velocity-based methods like Dynamo \cite{qiu2022mapping} use splicing information or time-resolved data to predict paths of differentiation from RNA data.

Despite the power and utility of these computational tools, most commonly employed methods for analyzing scRNA-seq data share certain limitations. Dimensional reduction is often one of the first steps in visualizing scRNA-seq data, but current methods highly distort cell-to-cell and cluster-to-cluster quantitative relationships \cite{chari_specious_2021}. Distorted embeddings inconsistently represent distances between individual data points and between cell types, making it difficult to interpret cell fate transitions. Similar issues arise in trajectory inference and RNA-velocity methods because local neighborhoods can also vary substantially with the choice of embedding. Another limitation of these methods is the lack of reproducibility inherent to stochastic machine learning methods like t-SNE and UMAP, which output different results on separate executions even when hyperparameters remain the same \cite{wattenberg_how_2016}. Many algorithms also perform feature selection (e.g., limiting their analysis to the most variable genes or highly-expressed genes) or require choosing parameters such as perplexity, making results even more sensitive to the inclusion or exclusion of different datasets in the analysis.

An ideal analysis algorithm for scRNA-seq data should possess certain desirable properties. First, the algorithm should incorporate explicit assumptions that can be kept consistent between analyses. Second, the algorithm should be deterministic, further enhancing reproducibility. Third, the axes in visualization should be interpretable, intuitive, and remain fixed across analysis and datasets. Fourth, the algorithm should assign a continuous value to cell identity instead of assigning discrete identity labels, thus enabling a more subtle and fine-grained analysis of cellular identity and developmental trajectories that reflect the continuous nature of biology. Finally, the algorithm should be easily able to incorporate datasets from multiple data sources, including the wealth of new single-cell atlases now being generated \cite{quake2022decade}.

Here, we present a new physics-inspired algorithm, single-cell Type Order Parameters (scTOP), that has all of these desired properties. The algorithm is based on the statistical physics concept of an order parameter -- a coarse-grained quantity that can distinguish between phases of matter. Arguably the most famous example of an order parameter is magnetization in a ferromagnetic system (i.e., a system that can spontaneously magnetize). In a ferromagnet, the magnetization measures the alignment of spins and can be used to distinguish magnetic and paramagnetic phases \cite{landau2013statistical,sethna2021statistical}. It is straightforward to generalize the idea of magnetization to more complex settings such as spin glasses and attractor neural networks \cite{amit1985spin}. In the context of epigenetic landscape models of cell identity, this takes the form of generalized magnetizations that measure how aligned a gene expression state is with each of the attractor basins for different cell fates \cite{lang_epigenetic_2014,pusuluri_cellular_2017}. These generalized magnetizations serve as natural order parameters for cellular identity and can be calculated directly from data \cite{lang_epigenetic_2014,pusuluri_cellular_2017, dame_thyroid_2017,ikonomou_vivo_2020}. The scTOP algorithm builds on previous work by developing methods for integrating scRNA-seq data, greatly expanding the power and utility of this approach.

The inputs to scTOP are the scRNA-seq measurements of query data as well as a reference basis containing the archetypal gene expression profiles for cell types of interest. The outputs of scTOP are the coordinates of the query data in cell type order parameter space (i.e., generalized magnetizations), with each output dimension serving as a measure of similarity between the query data and a distinct cell type in the reference basis. In this way, scTOP naturally projects gene expression deterministically onto a space with dimension equal to the number of cell types in the reference basis. These generalized coordinates can be used for both cell type identification and visualization. One appealing feature of scTOP is that the assumptions of the algorithm are explicit and are contained entirely in the choice of reference basis. This allows direct comparison across datasets and settings since the coordinates of and distances between data points do not change as long as one uses the same reference basis. One important limitation of scTOP worth noting is that it cannot be easily used to identify novel cell fates. The underlying reason for this is that scTOP assumes one already knows what cell types are of interest in the form of a reference basis and, for this reason, is less suited for exploratory work with undocumented cell types.

In this paper, we show that scTOP can be used to reliably assess cell identity and accurately classify cell type with high accuracy at the resolution of individual cells. We demonstrate this using mouse and human samples from lungs, brains, and other organs, including transplanted cells that were generated through the directed differentiation of pluripotent stem cells (PSCs). By analyzing lineage tracing data and embryonic development data, we also establish scTOP's capacity to quantify transitions between cell types, including bifurcations between closely related cell fates. We also show that scTOP can be used to quantitatively assess transcriptional similarity of engineered and natural cells by analyzing murine pulmonary transplantation experiments.

We have implemented scTOP in an easy-to-use Python package, available on the \href{https://pypi.org/project/scTOP/}{Python Package Index} and \href{https://github.com/Emergent-Behaviors-in-Biology/scTOP}{Github}. The code written for the analyses in this paper is available on \href{https://github.com/Emergent-Behaviors-in-Biology/scTOP-manuscript}{Github}.

\section{Methods}
\begin{figure}
	\centering
		\includegraphics[scale=0.8]{figs/scTOP manuscript fig1 - no aggregate.pdf}
	\caption{Steps involved in the scTOP algorithm. (a) Each cell in the query data is preprocessed independently of the other cells in the dataset. (b) scRNA-seq atlases, such as the Mouse Cell Atlas, are used to define the reference basis in the algorithm. (c) scTOP scores are the projections of query data onto the non-orthogonal subspace of cell types. Since there is no statistical fitting, no tuning parameters are involved. (d) The process of finding scTOP scores is shown in the grey section, and the blue section illustrates cell-type space. This space has as many dimensions as there are cell types in the reference basis. We can visualize this concept in 1, 2, or 3 dimensions, depending on which cell types are most relevant. In the 3-dimensional case, the shadows on the bounds of the scatter plot are included to better illustrate the position of points in 3-dimensional space.}
	\label{FIG:1}
\end{figure}

>>>>

\documentclass[aps,superscriptaddress, notitlepage,longbibliography]{revtex4-1}
\usepackage{hyperref}
\usepackage{amsmath}
\usepackage{physics}
\usepackage[pdftex]{graphicx}
\usepackage{xcolor}
\setlength {\marginparwidth }{2cm}
\hypersetup{
    colorlinks,
    citecolor=blue,
    filecolor=black,
    linkcolor=red,
    urlcolor=blue
}

@@

\title{scTOP: physics-inspired order parameters for cellular identification and visualization}

\author{Maria Yampolskaya}
\email{mariay@bu.edu}
\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
\author{Michael J Herriges}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{The Pulmonary Center and Department of Medicine, Boston University School of Medicine, Boston, MA, USA}
\author{Laertis Ikonomou}
\affiliation{Department of Oral Biology. University at Buffalo, The State University of New York, Buffalo, NY, USA}
\affiliation{Division of Pulmonary, Critical Care and Sleep Medicine, Department of Medicine, University at Buffalo, The State University of New York, Buffalo, NY, USA}
\author{Darrell N Kotton}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{The Pulmonary Center and Department of Medicine, Boston University School of Medicine, Boston, MA, USA}
\author{Pankaj Mehta}
\email{pankajm@bu.edu}
\affiliation{Department of Physics, Boston University, Boston, MA 02215, USA}
\affiliation{Center for Regenerative Medicine of Boston University and Boston Medical Center, Boston, MA, USA}
\affiliation{Faculty of Computing and Data Science, Boston University, Boston, MA 02215, USA}
\affiliation{Biological Design Center, Boston University, Boston, MA 02215, USA}

@@

\section*{Supplementary Information}
\beginsupplement

\subsection{Interpreting scTOP scores}\label{interpretation}
In theory, the attractor-network order parameters we have adapted for gene expression data range between negative one and positive one. An attractor state order parameter with a value of positive one means the system is perfectly aligned with that particular attractor state, and negative one means it is the exact opposite of that attractor state. Zero means that it is neither correlated nor anti-correlated with that attractor state.

In practice, scTOP scores generally do not reach positive one. scRNA-seq data contains significant numbers of dropouts, and these zeroes decrease the magnitude of the scTOP score. When an aggregate score is taken, the effect of dropouts is decreased and the scTOP score increases. This is important to keep in mind when interpreting scTOP scores; any score above 0.1 is notable, and since the data is so noisy even the best-case score may be well below 1. It's helpful to have a control population to compare with, to have a measure of the range of best-case scTOP scores. This is also why it's best to compare distributions of scores rather than scores for individual cells, since noise plays a non-negligible role.

scTOP scores are rarely negative. Negative values for scTOP scores correspond to cases where the sample gene expression profile is the opposite of the reference profile. For example, a score of -1 would indicate that all the genes which are highly-expressed in the reference are lowly-expressed in the sample, and vice versa. There are cases where the scores are low, negative values, indicating that the sample does not align well with the reference and that some genes which are highly- (lowly-) expressed in the reference are lowly- (highly-) expressed in the sample. From the data, it appears that it’s unlikely for gene networks to contain cell types which are opposites of one another. One possible reason is that some genes, like housekeeping genes, are similarly expressed across all cell types. Also, it’s possible that cells which expressed the exact opposite genes of any given cell type would not be able to survive.

\subsection{Comparison to other linear methods}
scTOP is mathematically and conceptually distinct from other linear methods such as Principal Component Analysis (PCA). PCA takes input data and finds the vectors in feature space along which the most sample-to-sample variation occurs. PCA is unsupervised and assumes nothing is known about the input data. The principal components need to be examined to determine their biological relevance. The major difference with scTOP is that scTOP takes input data and a reference basis of known cell types. The input data is decomposed onto the vectors defined by the reference basis instead of vectors corresponding to the most sample-to-sample variation. With scTOP, the resulting decomposition has a direct interpretation: alignment with each of the known cell types within the reference basis. It’s not possible to find the decomposition onto known cell types with PCA, because PCA doesn’t have the extra information about pre-defined cell type expression profiles.

To summarize, scTOP and PCA both transform the sample data to a different coordinate system, but there are major differences between the spaces that these methods project onto. The following table breaks down the differences between these spaces.

\begin{tabular}{ |p{3in}|p{3in}|} 
 \hline
 scTOP space & PCA space\\ 
 \hline
 \hline
 Defined by an input reference basis of known cell types & Calculated by PCA by finding the orthogonal vectors (principal components) that best explain the variation in data  \\ 
 \hline
 Can be non-orthogonal (e.g. can have highly-correlated cell types) & Orthogonal by definition  \\ 
 \hline
 Dimensions have clear biological interpretation: alignment with cell types & Dimensions need to be closely analyzed to be interpreted \\
 \hline
\end{tabular}

\subsection{Preprocessing} \label{preprocessing}
The raw scRNA-seq count data for each dataset was downloaded as a matrix of mRNA counts corresponding to genes and cells. An example of such a matrix is shown in step 0 of figure \ref{FIG:supp1}, where each row is a different gene, and each column is a different cell. Throughout this figure, the histogram to the right of each step shows the distribution of entry values for the first column (cell) in the matrix, and the histogram below the step shows the distribution of entry values for the second row (gene Rpl7) of the matrix. The histograms are presented as examples of how the distributions (for each cell and gene) of gene expressions change as the data is preprocessed.
